A thermodynamic view of dusty protoplanetary disks

### 2017-09-23 appendix.tex
additional edits in response to guillaume's comments
update to re-submitted version, unbolded

\subsection{Incompressible limit}

 We can set $\zeta = 0$ or consider $c_s^2\to \infty$ to
obtain the incompressible limit. Using $\tau_\mathrm{s} \equiv
\tstop/\left(1 - \tepsilon\right)$, we obtain: 
\begin{align}
\ii \tepsilon \tau_\mathrm{s}
  \sigma^4 - \sigma^3 - 
  \tau_\mathrm{s} \left[ \ii \tepsilon \kappa^2 + k_x F_r\left( 1 -
  \tepsilon\right)\right]\sigma^2 
  + \left(\frac{k_z\kappa}{|\bm{k}|}\right)^2\sigma + 
  k_x \tau_\mathrm{s} F_r\left(1 - 2\tepsilon\right)
  \left(\frac{k_z\kappa}{|\bm{k}|}\right)^2  = 0,\label{streaming_incompressible}
\end{align}
as derived by \cite{jacquet11} and \cite{laibe14} for the streaming
instability. 
If $|\sigma|/\OmK = O(\tau_\mathrm{s}\OmK)$ and
$\tau_\mathrm{s}\OmK\ll 1$, then the quartic term
is small and may be neglected. In that case we obtain the cubic
dispersion relation of \citet{youdin05a}.

@@

\subsubsection{Consistency check}\label{si_dust_rich_approx}
{\bf Here we check if the above explicit solutions are
  consistent with the assumptions used to obtain them.  
} 
Inserting the solutions for the most unstable mode 
%(Eq. \ref{si_dust_rich_optkx}---\ref{si_dust_rich_grow} with ) 
{\bf we find Eq. \ref{min_kz} becomes}
\begin{align}
  \fdust \gg \frac{12}{17}\simeq 0.7.
\end{align}
Since $\fdust \leq 1$, this requirement can only be marginally
satisfied. However, we find that this mostly give errors in the real
frequency (see Fig. \ref{si_compare_fig}), while growth rates are
still captured correctly. 
On the other hand, the assumption of Eq. \ref{max_kz} becomes the trivial
inequality \begin{align} 
  1\ll \frac{27}{4}\left(\frac{5}{3}\right)^3\simeq 30,
\end{align}
so Eq. \ref{max_kz} is satisfied, which justifies the approximate solution for 
$\mu$ in Eq. \ref{mu_explicit}.

{\bf Ultimately, the validity of these assumptions are justified
   a posteriori by comparison with the solution to the full equations
   in Fig. \ref{si_compare_fig}. 
}

>>>>

\subsection{Incompressible { gas} limit}

 We can set $\zeta = 0$ or consider $c_s^2\to \infty$ to
obtain the incompressible { gas} limit. Using $\tau_\mathrm{s} \equiv
\tstop/\left(1 - \tepsilon\right)$, we obtain: 
\begin{align}
\ii \tepsilon \tau_\mathrm{s}
  \sigma^4 - \sigma^3 - 
  \tau_\mathrm{s} \left[ \ii \tepsilon \kappa^2 + k_x F_r\left( 1 -
  \tepsilon\right)\right]\sigma^2 
  + \left(\frac{k_z\kappa}{|\bm{k}|}\right)^2\sigma + 
  k_x \tau_\mathrm{s} F_r\left(1 - 2\tepsilon\right)
  \left(\frac{k_z\kappa}{|\bm{k}|}\right)^2  = 0,\label{streaming_incompressible}
\end{align}
as derived by \cite{jacquet11} and \cite{laibe14} for the streaming
instability. 
If $|\sigma|/\OmK = O(\tau_\mathrm{s}\OmK)$ and
$\tau_\mathrm{s}\OmK\ll 1$, then the quartic term
is small and may be neglected. In that case we obtain the cubic
dispersion relation of \citet{youdin05a}.

@@

\subsubsection{Consistency check}\label{si_dust_rich_approx}
{ Here we check if the above explicit solutions are
  consistent with the assumptions used to obtain them.  
} 
Inserting the solutions for the most unstable mode 
%(Eq. \ref{si_dust_rich_optkx}---\ref{si_dust_rich_grow} with ) 
{ we find Eq. \ref{min_kz} becomes}
\begin{align}
  \fdust \gg \frac{12}{17}\simeq 0.7.
\end{align}
Since $\fdust \leq 1$, this requirement can only be marginally
satisfied. However, we find that this mostly give errors in the real
frequency (see Fig. \ref{si_compare_fig}), while growth rates are
still captured correctly. 
On the other hand, the assumption of Eq. \ref{max_kz} becomes the trivial
inequality \begin{align} 
  1\ll \frac{27}{4}\left(\frac{5}{3}\right)^3\simeq 30,
\end{align}
so Eq. \ref{max_kz} is satisfied, which justifies the approximate solution for 
$\mu$ in Eq. \ref{mu_explicit}.

{ Ultimately, the validity of these assumptions are justified
   a posteriori by comparison with the solution to the full equations
   in Fig. \ref{si_compare_fig}. 
}

### 2017-09-23 linear.tex
update to re-submitted version, unbolded

We assume the disk is approximately Keplerian so that
$\kappa\simeq\OmK$, and that the vertical gas distribution is   
Gaussian. In terms of the dust-to-gas
ratio $\epsilon$, Eq. \ref{dusty_solberg2}
becomes 
\begin{align}
  % &1 + \frac{\epsilon}{\left(1+
  %   \epsilon\right)^2}\left(h_\mathrm{g}^2\frac{\p\ln{\rhog}}{\p\ln{r}}
  % \frac{\p\ln{\epsilon}}{\p\ln{r}} + 
  % \frac{z^2}{\Htilde^2}\right)>0 \label{stability_est1},   \\ 
% &1 - \frac{\epsilon
%   h_\mathrm{g}^2}{\left(1+\epsilon\right)^2}
%   \frac{\p\ln{\epsilon}}{\p\ln{r}}\left(-\frac{\p\ln{\rhog}}{\p\ln{r}}+\frac{\Htilde^2}{H_\mathrm{g}^2}\frac{\p\ln{\epsilon}}{\p\ln{r}}\right)
%   > 0 \label{stability_est2},
&1 - \frac{\epsilon
  h_\mathrm{g}^2}{\left(1+\epsilon\right)^2}
  \frac{\p\ln{\epsilon}}{\p\ln{r}}\left(-\frac{\p\ln{P}}{\p\ln{r}}+\frac{\Htilde^2}{H_\mathrm{g}^2}\frac{\p\ln{\epsilon}}{\p\ln{r}}\right)
  > 0 \label{stability_est2},
\end{align}
for \emph{stability}. The {\bf first term in  brackets} is usually stabilizing in 
PPDs for reasons given in \S\ref{iso_perfect}. The {\bf second term in
brackets} is 
always destabilizing, but is small in smooth, thin disks with radial
gradients $O(1/r)$, $h_\mathrm{g}\ll 1$, 
provided that $H_\epsilon/\Hg$ is not large (e.g. some dust settling
has occurred). This means that typical dusty PPDs are stable to
axisymmetric perturbations, even for arbitrarily thin
dust layers.

@@

We can neglect pressure gradients in Eq. \ref{spm} 
if $r\p_r\ln{P}\sim O(1)$ and $H_\epsilon/\Hgas\gg
\sqrt{\epsilon}\hgas/2\left(1+\epsilon\right)$. For example, if $\epsilon\simeq 0.01$ and
$\hgas\simeq 0.05$, then we require $H_\epsilon/\Hg\gg
2\times10^{-3}$. This can be met if the dust is not well settled {\bf (e.g. due to a small amount of turbulence).}  
%this turbulence cannot be too big because the analysis itself does not have diffusion/turbulence 
%large diffusion is expected to stabilize 
Then instability requires 
\begin{align}\label{dust_edge}
\left|\frac{\p\ln{\epsilon}}{\p r}\right| \gtrsim
  \frac{1}{\Htilde}\frac{\left(1+\epsilon\right)}{\sqrt{\epsilon}}.
%\simeq
 % \frac{1}{\sqrt{\epsilon}H_\epsilon}, 
\end{align}
%where the second equality applies to $\epsilon\ll 1$. 
That is, if the radial lengthscale of the dust-to-gas ratio 
is much less its vertical lengthscale, $L_\epsilon\ll O(\Htilde)$, 
then the system is  unstable.  
Taking $\epsilon\sim 0.01$,  we find that 
for thin dust layers with $H_\epsilon \simeq \Hd\ll \Hg$,  instability
requires $L_\epsilon\ll O(0.1\Hg)$, i.e. the dust-to-gas ratio must 
vary on an extremely short radial lengthscale. This might be achieved, for
example, at sharp edges associated with gaps opened by giant planets.

>>>>

We assume the disk is approximately Keplerian so that
$\kappa\simeq\OmK$, and that the vertical gas distribution is   
Gaussian. In terms of the dust-to-gas
ratio $\epsilon$, Eq. \ref{dusty_solberg2}
becomes 
\begin{align}
  % &1 + \frac{\epsilon}{\left(1+
  %   \epsilon\right)^2}\left(h_\mathrm{g}^2\frac{\p\ln{\rhog}}{\p\ln{r}}
  % \frac{\p\ln{\epsilon}}{\p\ln{r}} + 
  % \frac{z^2}{\Htilde^2}\right)>0 \label{stability_est1},   \\ 
% &1 - \frac{\epsilon
%   h_\mathrm{g}^2}{\left(1+\epsilon\right)^2}
%   \frac{\p\ln{\epsilon}}{\p\ln{r}}\left(-\frac{\p\ln{\rhog}}{\p\ln{r}}+\frac{\Htilde^2}{H_\mathrm{g}^2}\frac{\p\ln{\epsilon}}{\p\ln{r}}\right)
%   > 0 \label{stability_est2},
&1 - \frac{\epsilon
  h_\mathrm{g}^2}{\left(1+\epsilon\right)^2}
  \frac{\p\ln{\epsilon}}{\p\ln{r}}\left(-\frac{\p\ln{P}}{\p\ln{r}}+\frac{\Htilde^2}{H_\mathrm{g}^2}\frac{\p\ln{\epsilon}}{\p\ln{r}}\right)
  > 0 \label{stability_est2},
\end{align}
for \emph{stability}. The { first term in  brackets} is usually stabilizing in 
PPDs for reasons given in \S\ref{iso_perfect}. The { second term in
brackets} is 
always destabilizing, but is small in smooth, thin disks with radial
gradients $O(1/r)$, $h_\mathrm{g}\ll 1$, 
provided that $H_\epsilon/\Hg$ is not large (e.g. some dust settling
has occurred). This means that typical dusty PPDs are stable to
axisymmetric perturbations, even for arbitrarily thin
dust layers.

@@

We can neglect pressure gradients in Eq. \ref{spm} 
if $r\p_r\ln{P}\sim O(1)$ and $H_\epsilon/\Hgas\gg
\sqrt{\epsilon}\hgas/2\left(1+\epsilon\right)$. For example, if $\epsilon\simeq 0.01$ and
$\hgas\simeq 0.05$, then we require $H_\epsilon/\Hg\gg
2\times10^{-3}$. This can be met if the dust is not well settled { (e.g. due to a small amount of turbulence).}  
%this turbulence cannot be too big because the analysis itself does not have diffusion/turbulence 
%large diffusion is expected to stabilize 
Then instability requires 
\begin{align}\label{dust_edge}
\left|\frac{\p\ln{\epsilon}}{\p r}\right| \gtrsim
  \frac{1}{\Htilde}\frac{\left(1+\epsilon\right)}{\sqrt{\epsilon}}.
%\simeq
 % \frac{1}{\sqrt{\epsilon}H_\epsilon}, 
\end{align}
%where the second equality applies to $\epsilon\ll 1$. 
That is, if the radial lengthscale of the dust-to-gas ratio 
is much less its vertical lengthscale, $L_\epsilon\ll O(\Htilde)$, 
then the system is  unstable.  
Taking $\epsilon\sim 0.01$,  we find that 
for thin dust layers with $H_\epsilon \simeq \Hd\ll \Hg$,  instability
requires $L_\epsilon\ll O(0.1\Hg)$, i.e. the dust-to-gas ratio must 
vary on an extremely short radial lengthscale. This might be achieved, for
example, at sharp edges associated with gaps opened by giant planets.

### 2017-09-23 model.tex
update to re-submitted version, unbolded

\subsection{Thermodynamics of dust-drag instabilities}\label{dust_work}
The physical interpretation of \S\ref{grow_osc} is here derived in
detail. Consider constant $c_s$ but $\tstop\neq0$ (so $\delta\mathcal{C}
\neq 0$). Eq. \ref{int_rel} indicates that $\sigma^2$ is generally complex, {\bf unless
the second integral on the right-hand-side is real. Thus } 
we may have growing oscillations, or overstability, due to dust-gas
friction. This is seen by taking the imaginary part of
Eq. \ref{int_rel} 
\begin{align}
%  s = \frac{\imag\int \left(\nabla\cdot\dd\bm{v}^*\right)\dd\mathcal{C}dV}{2\omega\int\rho\left(|\dd
%    v_r|^2 + |\dd v_z|^2\right)dV}, \label{thermal_instability}
  s = \frac{\imag\int \left(\nabla\cdot\dd\bm{v}^*\right)\dd\mathcal{C}dV}{2\omega\mathcal{I}^2}, \label{thermal_instability}
\end{align} 
assuming $\omega\neq0$. %This represents overstability due to dust-gas friction. 
Since drag ($\delta \mathcal{C}$) appears as a source term in our
effective energy equation, the quantity 
$\imag\left[\left(\nabla\cdot\dd\bm{v}^*\right)\dd\mathcal{C}\right]$
represent correlations between compression/expansion and  
heating/cooling.

@@

The physical interpretation of Eq. \ref{thermal_instability} is that 
work done by pressure forces in the dusty gas leads to growth {\bf ($s>0$) or decay ($s<0$)} 
%{\bf work done by pressure forces in the dusty gas modifies
in oscillation amplitudes. 
To demonstrate this, we calculate the average
work done assuming periodic oscillations, and show that if the average
work done is positive, then the oscillation amplitude would actually grow.

@@

{\bf Here we consider Lagragian perturbations such that 
  \begin{align}
 P \to P + \real\left(\Delta P e^{-\ii\sigma t}\right)
 \end{align}
is the pressure following a fluid element of the mixture (and similarly for $\rho$). 
 The Lagrangian perturbation $\Delta$ of a variable $X$ is $\Delta X = \delta X + \bm{\xi}\cdot\nabla X$
 and $\bm{\xi}$ is the Lagrangian displacement, so $\xi_{x,z} =  \ii\dd
 v_{x,z}/\sigma$. Inserting the above pressure and density fields
 into Eq. \ref{work_def}, and noting that only products of
 perturbations contribute to $\mathcal{W}$ after time-averaging, we
 find for periodic oscillations (time dependence $e^{\ii\omega t}$ and
 real $\omega$) that:  
}
\begin{align}
  \mathcal{W} = - \frac{\omega}{2} \int \imag\left(\Delta P
  \frac{\Delta \rho^*}{\rho}\right)dV,\label{work_real}
\end{align}
Eq. \ref{work_real} show that a phase difference between gas pressure and
the total density leads to work done
($\mathcal{W}\neq0$).

@@

The above discussion applies to any single fluid with a pressure and
density. In the case of interest --- dusty gas --- the work done 
is attributed to finite dust-gas drag. 
%For dusty gas, the work done during oscillations 
 %is attributed to finite dust-gas drag. 
The relative 
drift between gas and dust{\bf, which only exists if $\tstop\nabla P \neq 0$ (Eq. \ref{term_vel_approx}),} causes a phase difference between the two
components, and hence between pressure and total density.

>>>>

\subsection{Thermodynamics of dust-drag instabilities}\label{dust_work}
The physical interpretation of \S\ref{grow_osc} is here derived in
detail. Consider constant $c_s$ but $\tstop\neq0$ (so $\delta\mathcal{C}
\neq 0$). Eq. \ref{int_rel} indicates that $\sigma^2$ is generally complex, { unless
the second integral on the right-hand-side is real. Thus } 
we may have growing oscillations, or overstability, due to dust-gas
friction. This is seen by taking the imaginary part of
Eq. \ref{int_rel} 
\begin{align}
%  s = \frac{\imag\int \left(\nabla\cdot\dd\bm{v}^*\right)\dd\mathcal{C}dV}{2\omega\int\rho\left(|\dd
%    v_r|^2 + |\dd v_z|^2\right)dV}, \label{thermal_instability}
  s = \frac{\imag\int \left(\nabla\cdot\dd\bm{v}^*\right)\dd\mathcal{C}dV}{2\omega\mathcal{I}^2}, \label{thermal_instability}
\end{align} 
assuming $\omega\neq0$. %This represents overstability due to dust-gas friction. 
Since drag ($\delta \mathcal{C}$) appears as a source term in our
effective energy equation, the quantity 
$\imag\left[\left(\nabla\cdot\dd\bm{v}^*\right)\dd\mathcal{C}\right]$
represent correlations between compression/expansion and  
heating/cooling.

@@

The physical interpretation of Eq. \ref{thermal_instability} is that 
work done by pressure forces in the dusty gas leads to growth { ($s>0$) or decay ($s<0$)} 
%{\bf work done by pressure forces in the dusty gas modifies
in oscillation amplitudes. 
To demonstrate this, we calculate the average
work done assuming periodic oscillations, and show that if the average
work done is positive, then the oscillation amplitude would actually grow.

@@

{ Here we consider Lagragian perturbations such that 
  \begin{align}
 P \to P + \real\left(\Delta P e^{-\ii\sigma t}\right)
 \end{align}
is the pressure following a fluid element of the mixture (and similarly for $\rho$). 
 The Lagrangian perturbation $\Delta$ of a variable $X$ is $\Delta X = \delta X + \bm{\xi}\cdot\nabla X$
 and $\bm{\xi}$ is the Lagrangian displacement, so $\xi_{x,z} =  \ii\dd
 v_{x,z}/\sigma$. Inserting the above pressure and density fields
 into Eq. \ref{work_def}, and noting that only products of
 perturbations contribute to $\mathcal{W}$ after time-averaging, we
 find for periodic oscillations (time dependence $e^{\ii\omega t}$ and
 real $\omega$) that:  
}
\begin{align}
  \mathcal{W} = - \frac{\omega}{2} \int \imag\left(\Delta P
  \frac{\Delta \rho^*}{\rho}\right)dV,\label{work_real}
\end{align}
Eq. \ref{work_real} show that a phase difference between gas pressure and
the total density leads to work done
($\mathcal{W}\neq0$).

@@

The above discussion applies to any single fluid with a pressure and
density. In the case of interest --- dusty gas --- the work done 
is attributed to finite dust-gas drag. 
%For dusty gas, the work done during oscillations 
 %is attributed to finite dust-gas drag. 
The relative 
drift between gas and dust{, which only exists if $\tstop\nabla P \neq 0$ (Eq. \ref{term_vel_approx}),} causes a phase difference between the two
components, and hence between pressure and total density.

### 2017-09-23 paper.tex
updated to revision 1 submitted version
update to re-submitted version, unbolded

This paper is organized as follows.  
In \S\ref{grow_osc} we describe a key insight of our study, that
pressure (i.e.\ ``PdV")  work {\bf can} explain the basis of several
instabilities, including SI and VSI, in terms of 
pressure-density phase lags. {\bf This motivates us to seek an 
  analogy between dust-gas mixtures and pure gas.}    
We develop our formalism in \S\ref{setup} by transforming the
two fluid equations of a tightly-coupled dusty gas into single fluid
hydrodynamic equations with a special cooling function.  Here we also
define the effective entropy and buoyancy of our model fluid. 
In \S\ref{limits} we discuss general stability properties of dusty
disks. We then analyze explicit disk models 
\S\ref{linear_problem}, were we show that radial dust edges are unstable,
and revisit the SGI and SI in the tight 
coupling regime. In \S\ref{results} we 
extend the VSI to dusty disks. We   
discuss future applications of the dusty/adiabatic gas equivalence in 
\S\ref{discussion} before summarizing in \S\ref{summary}.

@@

\section{Growing oscillations by doing work}\label{grow_osc}
{\bf
In this section we provide physical arguments to highlight the
similarity between single phase fluids and dust-gas mixtures. It is
well-known that in the limit of perfect coupling the addition of dust
increases the gas inertia but not pressure. One can then regard 
dusty gas as a single fluid with a reduced sound-speed. Here, we
argue that when dust-gas coupling is imperfect, there exists another
similarity with pure gas; related the phase of pressure and density
evolution. Thus it is still possible to regard
partially-coupled dusty-gas as a single fluid. The mathematical
description of this section is given in \S\ref{dust_work}.   
}
%dust_work

@@

Another possibility, as annotated in Fig. \ref{pdv_cartoon}, is a dusty gas. The
relevant density here is the total density $\rho = \rhod + \rhog$, 
but the fluid pressure $P$ is due to gas only. If dust were 
perfectly-coupled to the gas, then $P$ and $\rho$ would be in phase, 
and there is no work done. However, for finite dust-gas drag, $\rhod$ and
$\rhog$ are not necessarily in phase, because dust particles can
drift relative to the gas{\bf, i.e. the dust-to-gas ratio evolves in time}. 
If this causes $P$ to lag 
$\rho$, then the positive work done would lead to growing
oscillations. Indeed, we show that this is true for the streaming
instability.

>>>>

This paper is organized as follows.  
In \S\ref{grow_osc} we describe a key insight of our study, that
pressure (i.e.\ ``PdV")  work {can} explain the basis of several
instabilities, including SI and VSI, in terms of 
pressure-density phase lags. { This motivates us to seek an 
  analogy between dust-gas mixtures and pure gas.}    
We develop our formalism in \S\ref{setup} by transforming the
two fluid equations of a tightly-coupled dusty gas into single fluid
hydrodynamic equations with a special cooling function.  Here we also
define the effective entropy and buoyancy of our model fluid. 
In \S\ref{limits} we discuss general stability properties of dusty
disks. We then analyze explicit disk models 
\S\ref{linear_problem}, were we show that radial dust edges are unstable,
and revisit the SGI and SI in the tight 
coupling regime. In \S\ref{results} we 
extend the VSI to dusty disks. We   
discuss future applications of the dusty/adiabatic gas equivalence in 
\S\ref{discussion} before summarizing in \S\ref{summary}.

@@

\section{Growing oscillations by doing work}\label{grow_osc}
{
In this section we provide physical arguments to highlight the
similarity between single phase fluids and dust-gas mixtures. The mathematical
description of this section is deferred to \S\ref{dust_work}.

It is
well-known that in the limit of perfect coupling the addition of dust
increases the gas inertia but not pressure. One can then regard 
dusty gas as a single fluid with a reduced sound-speed. Here, we
argue that when dust-gas coupling is imperfect, there exists another
similarity with pure gas; related the phase of pressure and density
evolution. Thus it is still possible to regard
partially-coupled dusty-gas as a single fluid.  
}
%dust_work

@@

Another possibility, as annotated in Fig. \ref{pdv_cartoon}, is a dusty gas. The
relevant density here is the total density $\rho = \rhod + \rhog$, 
but the fluid pressure $P$ is due to gas only. If dust were 
perfectly-coupled to the gas, then $P$ and $\rho$ would be in phase, 
and there is no work done. However, for finite dust-gas drag, $\rhod$ and
$\rhog$ are not necessarily in phase, because dust particles can
drift relative to the gas{, i.e. the dust-to-gas ratio evolves in time}. 
If this causes $P$ to lag 
$\rho$, then the positive work done would lead to growing
oscillations. Indeed, we show that this is true for the streaming
instability.

### 2017-09-23 result.tex
update to re-submitted version, unbolded

Fig. \ref{compare_vshear_varHd} shows that a radially-varying
dust-to-gas ratio increases the magnitude of the vertical shear rate
away from the midplane (see also Eq. \ref{vshear2}). Thus we {\bf
  typically} find higher VSI growth rates, as shown in
Fig. \ref{vsi_dust_varHd}. 
Surfaces modes, {\bf appearing at high $k_x$}, are more effectively
enhanced by the additional vertical shear induced by
$\p_r\epsilon$. {\bf Low frequency body modes with small
  $k_x$ are more affected by radial variations in the 
  dust-to-gas ratio than high-$k_x$, high-frequency body modes. }

@@

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_eigenvals_varHd} 
  \caption{Unstable VSI modes in the disk models of
    Fig. \protect\ref{compare_vshear_varHd}. {\bf Diamonds (asterisks) 
    are mode frequencies for a disk with radially uniform (varying)
    dust-to-gas ratio. The radial wavenumbers are $k_xH_g = 10$
    (green),$k_xH_g = 30$ (black) and $k_xH_g = 100$ (orange). }
    \label{vsi_dust_varHd}
    }
\end{figure}

@@

In \S\ref{iso_perfect} we found that for strictly isothermal
disks, a vertically-uniform dust-to-gas ratio can be unstable
if $d\epsilon/dr\neq0$.   
{\bf
To demonstrate this numerically we set $q=0$ and $H_\mathrm{d}$ such
that $H_\epsilon=10^3\Hg$. We let $\rhod/\rhog \propto r^{-d}$
with different power indices $d$. The metalicity is fixed to $Z=0.03$.  
In these cases vertical shear $\p_z\Omega$ is 
attributed to the radially-varying dust-to-gas ratio (see
Eq. \ref{vshear2}).    
}

{\bf We show unstable modes in 
Fig. \ref{vert_mixed_modes}.} 
%In order to obtain appreciable growth   
%rates, we consider super-Solar metalicity $Z=0.03$ and high
%radial wavenumber $k_x\Hg=1800$. 
As expected from the discussion in
\S\ref{iso_perfect}, the disk admits purely growing modes with
$\omega=0$. This is distinct from the growing oscillations associated
with classic VSI discussed above.

{\bf We find in smooth disks large 
  $k_x$ is needed for appreciable growth rates, e.g. $k_xH_g=1800$
  with $\rhod/\rhog\propto r^{-1}$. 
  Local shearing box simulations may 
  be required to study the non-linear evolution of such short 
  wavelengths \citep[e.g.][]{bai10b,yang16}. Alternatively, as shown in Fig. \ref{vert_mixed_modes},
  a rapidly varying dust-to-gas ratio permits dynamical instability at
  longer radial wavelengths, which might be resolvable in global disk 
  simulations. 
}

>>>>

Fig. \ref{compare_vshear_varHd} shows that a radially-varying
dust-to-gas ratio increases the magnitude of the vertical shear rate
away from the midplane (see also Eq. \ref{vshear2}). Thus we {
  typically} find higher VSI growth rates, as shown in
Fig. \ref{vsi_dust_varHd}. 
Surfaces modes, { appearing at high $k_x$}, are more effectively
enhanced by the additional vertical shear induced by
$\p_r\epsilon$. { Low frequency body modes with small
  $k_x$ are more affected by radial variations in the 
  dust-to-gas ratio than high-$k_x$, high-frequency body modes. }

@@

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_eigenvals_varHd} 
  \caption{Unstable VSI modes in the disk models of
    Fig. \protect\ref{compare_vshear_varHd}. { Diamonds (asterisks) 
    are mode frequencies for a disk with radially uniform (varying)
    dust-to-gas ratio. The radial wavenumbers are $k_xH_g = 10$
    (green),$k_xH_g = 30$ (black) and $k_xH_g = 100$ (orange). }
    \label{vsi_dust_varHd}
    }
\end{figure}

@@

In \S\ref{iso_perfect} we found that for strictly isothermal
disks, a vertically-uniform dust-to-gas ratio can be unstable
if $d\epsilon/dr\neq0$.   
{
To demonstrate this numerically we set $q=0$ and $H_\mathrm{d}$ such
that $H_\epsilon=10^3\Hg$. We let $\rhod/\rhog \propto r^{-d}$
with different power indices $d$. The metalicity is fixed to $Z=0.03$.  
In these cases vertical shear $\p_z\Omega$ is 
attributed to the radially-varying dust-to-gas ratio (see
Eq. \ref{vshear2}).    
}

{ We show unstable modes in 
Fig. \ref{vert_mixed_modes}.} 
%In order to obtain appreciable growth   
%rates, we consider super-Solar metalicity $Z=0.03$ and high
%radial wavenumber $k_x\Hg=1800$. 
As expected from the discussion in
\S\ref{iso_perfect}, the disk admits purely growing modes with
$\omega=0$. This is distinct from the growing oscillations associated
with classic VSI discussed above.

{ We find in smooth disks large 
  $k_x$ is needed for appreciable growth rates, e.g. $k_xH_g=1800$
  with $\rhod/\rhog\propto r^{-1}$. 
  Local shearing box simulations may 
  be required to study the non-linear evolution of such short 
  wavelengths \citep[e.g.][]{bai10b,yang16}. Alternatively, as shown in Fig. \ref{vert_mixed_modes},
  a rapidly varying dust-to-gas ratio permits dynamical instability at
  longer radial wavelengths, which might be resolvable in global disk 
  simulations. 
}
